---
execute:
  echo: false
  message: false
  warning: false
---

<!-- I am using tidycencus and R. Give me code to download info about the percentage of of people with bachelor degrees in census tracts in LA
--> 

```{r}
# library(tidycensus)
# library(tidyverse)
# options(tidyverse.progress_bar = FALSE)
# options(tidycensus.progress_bar = FALSE)
# Sys.setenv(VROOM_SHOW_PROGRESS = "false")
# options(cli.progress_show_after = Inf)
# options(tigris_use_cache = TRUE)
# options(tigris_progress = FALSE)
```

```{r}

#| cache: TRUE
#| message:
# View educational attainment variables if needed
# v <- load_variables(2022, "acs5", cache = TRUE)
# View(v %>% filter(str_detect(name, "B15003")))

# edu_data <- get_acs(
  # geography = "tract",
  # variables = c(total = "B15003_001", bachelor_up = "B15003_022"),
  # state = "CA",
  # county = "Los Angeles",
  # year = 2022,
  # geometry = TRUE,
  # survey = "acs5"
# )

# Reshape and calculate percentage
# la_edu_percent <- edu_data %>%
  # select(GEOID, variable, estimate, geometry) %>%
  # pivot_wider(names_from = variable, values_from = estimate) %>%
  # mutate(percent_bachelor_or_higher = 100 * bachelor_up / total)

# View result
#glimpse(la_edu_percent)

```
<!--
here is my data make an interesting interactive plot using leaflet in R
Los Angeles Education
Rows: 2,498
Columns: 5
$ GEOID                      <chr> "06037670413", "06037650101", "06037620303"…
$ geometry                   <MULTIPOLYGON [°]> MULTIPOLYGON (((-118.4037 3...…
$ total                      <dbl> 3311, 4228, 3207, 3812, 3522, 3637, 2441, 2…
$ bachelor_up                <dbl> 1264, 1939, 1088, 1393, 827, 815, 197, 647,…
$ percent_bachelor_or_higher <dbl> 38.175778, 45.860927, 33.925787, 36.542497,…
-->

```{r}
library(tidyverse)
# load the data from rds file 
la_edu_percent <- read_rds("la_edu_percent.rds")
```

```{r}
library(tidycensus)


# Get tract names and county info for LA County
tract_metadata <- get_acs(
  geography = "tract",
  variables = "B01001_001",  # total population
  state = "CA",
  county = "Los Angeles",
  year = 2022,
  geometry = FALSE
) |>
  select(GEOID, NAME)

# Example NAME: "Census Tract 203.03, Los Angeles County, California"

# Join with your data
la_edu_percent <- la_edu_percent |>
  left_join(tract_metadata, by = "GEOID")

```


```{r}
#| message: false
# Load libraries
library(leaflet)
library(sf)
library(tidyverse)
library(RColorBrewer)



# Assuming your data is already loaded as `la_edu`
# and `geometry` is an sf column
# Make sure it's an sf object
la_edu <- st_as_sf(la_edu_percent)

# Create a color palette based on percent_bachelor_or_higher
pal <- colorNumeric(
  palette = "YlGnBu",
  domain = la_edu$percent_bachelor_or_higher
)

# Create labels for tooltips
labels <- sprintf(
  "<strong>%s</strong><br/>
   <strong>Total Pop 25+:</strong> %s<br/>
   <strong>With Bachelor's+:</strong> %s<br/>
   <strong>%% Bachelor's or Higher:</strong> %.1f%%",
  la_edu$NAME,
  la_edu$total,
  la_edu$bachelor_up,
  la_edu$percent_bachelor_or_higher
) %>% lapply(htmltools::HTML)


# Build leaflet map
leaflet(la_edu) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(percent_bachelor_or_higher),
    weight = 0.5,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "13px",
      direction = "auto")
  ) %>%
  addLegend(
    pal = pal,
    values = ~percent_bachelor_or_higher,
    opacity = 0.7,
    title = "% Bachelor's or Higher",
    position = "bottomright"
  ) |>
  # Add title as overlay control
  addControl(
    html = "<div style='text-align:center; font-size:18px; font-weight:bold;'>Bachelor's Degree Rates in Los Angeles County</div>",
    position = "topright"
  ) |>
  # Add subtitle
  addControl(
    html = "<div style='text-align:center; font-size:14px;'>% of Adults Aged 25+ with at Least a Bachelor's Degree (ACS 2022)</div>",
    position = "topright"
  ) |>
  # Add source
  addControl(
    html = "<div style='text-align:center; font-size:11px; color:gray;'>Source: U.S. Census Bureau, ACS 5-Year Estimates</div>",
    position = "bottomleft"
  )

```

This interactive map visualizes the percentage of adults aged 25 and over with a bachelor's degree or higher across census tracts in Los Angeles County. Using color gradients, it highlights spatial variation in educational attainment, with darker shades indicating higher percentages. Hovering over each tract reveals key details, including the total population, number with a bachelor's degree or more, and the exact percentage. This visualization allows for easy identification of geographic disparities in educational attainment and can support data-driven decisions in policy, planning, and community development.
